**The Heart can Lie: The Role of Interoception and Theory of Mind in Deception**

### 2022-11-15 paper/manuscript.Rmd
measures

---
title: '**Does the Heart Lies: The Role of Interoception and Theory of Mind in Deception**'
shorttitle        : "TITLE"
author:

@@

Besides paying to attention to the person we lie to, trying to gauge whether they believe us, it is also possible that some attention is directed inwards: monitoring our own body and its reactions (e.g., cardiac activity and its related changes such as blushing), susceptible to be used as cues to infer our real intent. This begs the question of the potential role of interoceptive abilities in deception ability. Broadly defined as one's sensitivity to their own internal signals and bodily states [@murphy2019classifying; @chen2021emerging; @weiss2014interaction], @garfinkel2015knowing's conceptualizes interoception as a three-dimensional framework comprising three distinct facets, namely, *interoceptive accuracy* - the objective ability to monitor internal bodily signals; *interoceptive sensibility* - the subjective confidence in one's interoceptive accuracy; and *interoceptive awareness* - one's ability to identify and appropriately respond to their perceived interoceptive states. Interoception has increasingly been tied to subjective perceptual experiences [@connell2018interoception; @seth2012interoceptive], as well as individual differences in executive functions, emotional processing and decision-making [@barrett2015interoceptive; @petzschner2021computational; @murphy2019classifying]. Although little research interest has been conferred to investigating the relationship between interoception and deceptive ability *per se*, previous decision-making studies have demonstrated a negative correlation between interoceptive awareness, a metacognitive dimension of interoception, and one's likelihood to make risky decisions [@dunn2010listening; @furman2013interoceptive]. In fact, this is consistent with the somatic marker hypothesis, which posits an association between interoception and a propensity towards making rational decisions [@damasio1996somatic].

@@

The aim of the present study was to explore the contribution of ToM and interoception abilities on individuals' deceptive skills, as indicated by their lying confidence, physiological arousal and response time. To this end, we designed a directed-lying paradigm with 2 conditions differing in the nature of their feedback cues. The *Interrogation* condition was designed to emphasize (and preferentially mobilize) ToM-related mechanims, whereas the *Polygraph* condition was designed to emphasize interoceptive mechanisms. We expected ToM and interoception to positively predict lying ability (i.e., higher lie confidence, shorter response time and lower physiological arousal), in particular in the *Polygraph* and the *Interrogation* condition, respectively.

@@

30 participants were recruited through  posters, flyers, and online social media platforms. Inclusion criteria included residing in Singapore, and an absence of current or history of neurological, psychiatric, or cardiovascular disorders. 4 participants were excluded as a result of a technical error that led to their data not being recorded. The final sample consists 26 participants (Mean age = 20.9, SD = 2.0, range:[18, 25], Sex: 65.4% women, 34.6% men, 0.00% non-binary). The heart rate of one participant and response time of one participant were excluded from further analysis due to extreme outlying values [**PLEASE HELP ME CONFIRM**].

This pre-registered study (**OSF?**) was approved by the NTU Institutional Review Board (NTU IRB-2018-10-029)(**TBC**) and all procedures performed were in accordance with the ethical standards of the institutional board and with the 1964 Helsinki Declaration. All participants provided their informed consent prior to participation and were awarded with academic credit upon completion of the study.

@@

The *Basic Empathy Scale (BES)*, a 20-item, 5-point Likert scale (Strongly Disagree - Strongly Agree) that measures cognitive and affective empathy [@jolliffe2006development] was administered. Although TOM and empathy are regarded as distinct psychological constructs, previous research findings point to them being closely related [@gallant2020developing; @sebastian2012neural]. Specifically, empathy is often thought to be an integral component in the affective dimension of TOM (i.e., the ability to infer what someone else is feeling) [@shamay2010role]. The *Yoni Task*, which comprises 3 dimensions, namely cognitive, physical and affective TOM, was also administered to assess participants' objective TOM ability [@shamay2007dissociable].

### Interoceptive Abilities

To assess participants' interoceptive ability, participants completed a *Heartbeat Counting Task (HCT)* [@schandry1981heart] while having their actual heartbeats recorded. During the HCT task, participants were instructed to count the number of heartbeats they felt over 5 trials with varying time intervals (20s, 25s, 30s, 35s, 40s), the sequence of which was randomized. Objective interoceptive accuracy (*HCT Accuracy*) was then computed based on the following formula ($\HCT_{Acc} = 1-|actual-count|/(actual + count)/2)$) [**lol pls help me write this in proper syntax**). To assess their interoceptive sensibility (*HCT Confidence*), participants were asked to rate, on a visual analog scale, how confident they were of their heartbeat counting accuracy at the end of each interval. Interoceptive awareness (*HCT Awareness*) was indexed by the correspondence between one's objective interoceptive accuracy and their confidence ratings.

The *MAIA-2* [@mehling2012multidimensional], a 32-item questionnaire which measures 8 dimensions of interoception (Noticing, Not-Distracting, Not-Worrying, Attention, Regulation, Emotional Awareness, Self-Regulation, Body Listening, and Trust) using a 5-point Likert scale (Never-Always) was also administered.

### Lie Profile

To evaluate participants' tendency to lie in their everyday lives, we used the *LIE Scale*, a 16-item questionnaire that assesses 4 dispositional lying dimensions (Ability, Negativity, Contextuality and Frequency) on a visual analog scale (Disagree-Agree) [@makowski2021structure].

### Demographic Information

We further collected demographic information including age, ethnicity, nationality, handedness (for application of EDA electrodes), religion, duration lived in Singapore as well as highest education qualification achieved (or in the midst of achieving).

### Deception Task Paradigm

We designed a directed-lying bi-modal feedback task (the *Deception Task*) consisting 2 conditions - namely, the *Interrogation* condition and the *Polygraph* condition.

In the *Interrogation* condition, participants were given social feedback in the form of a video stream with an examiner. Notably, participants were informed that the examiner would be evaluating the truthfulness of their responses by observing them through the live video feed. In actuality, the video was a pre-recorded video clip of a confederate, and the same video-clip was presented to all participants.

In the *Polygraph* condition, participants were given biofeedback in the form of an electrocardiogram (ECG) displayed onscreen instead. Specifically, participants were informed that they were being shown live feedback of their own heartbeats when in fact, the displayed ECG was that of a pre-recorded video clip. Similar to the *Interrogation* condition, all participants were shown the same pre-recorded video clip of the ECG.

The sequence of each task trial was the same for both conditions. Using a computer, participants were first shown a question, either phrased directly (*"What is your favourite sport?*), or indirectly (*"Is your favourite sport Hockey?"*), with accompanying instructions below that either asked participants to verbally lie or tell the truth. Following a connecting screen (0.7 - 1.5s), participants were presented with either social (*Interrogation* condition) or biofeedback (*Polygraph* condition) that was displayed for a maximum of 10s or until they made their accorded response. A disconnecting screen was then displayed (1.5-2.5s), following which participants were asked to rate their confidence in how convincing they perceived their answer to be.

@@

In the first phase of the study, participants answered a brief demographic survey before being administered a battery of questionnaires (*BES*, *MAIA-2*, *Lie Scale*) and cognitive-behavioural tasks (*HCT*, *Yoni Task*). This series of tasks and questionnaires were administered in a randomized order [**TBC**] in order to control for order effects. In the second phase of the study, the *Deception Task* was conducted using PsychoPy [@peirce2019psychopy2]. Each participant was administered both conditions of the *Deception Task*, the order sequence of which was randomized. Each task condition comprised 40 trials, of which participants were instructed to lie in half of the trials (20 **Truth** trials; 20 **Lie** trials). In total, each participant underwent 80 *Deception Task* trials.

Physiological recording devices were prepared prior to beginning the experimental paradigm. Using three ECG electrodes placed according to a modified Lead II configuration [@takuma1995alternative], cardiovascular activity, sampled at 1000Hz (**TBC**) was recorded via the BioPac MP160 system (BioPac Systems Inc., USA) and the AcqKnowledge 5.0 software.

>>>>

---
title: '**The Heart can Lie: The Role of Interoception and Theory of Mind in Deception**'
shorttitle        : "Deception, Interoception, and ToM"
author:

@@

Besides paying to attention to the person we lie to, trying to gauge whether they believe us, it is also possible that some attention is directed inwards: monitoring our own body and its reactions (e.g., cardiac activity and its related changes such as blushing), susceptible to be used as cues to infer our real intent. This begs the question of the potential role of interoceptive abilities in deception ability. Broadly defined as one's sensitivity to their own internal signals and bodily states [@murphy2019classifying; @chen2021emerging; @weiss2014interaction], @garfinkel2015knowing's conceptualizes interoception as a three-dimensional framework comprising three distinct facets, namely, interoceptive accuracy - the objective ability to monitor internal bodily signals; interoceptive sensibility - the subjective confidence in one's interoceptive accuracy; and interoceptive awareness - one's ability to identify and appropriately respond to their perceived interoceptive states. Interoception has increasingly been tied to subjective perceptual experiences [@connell2018interoception; @seth2012interoceptive], as well as individual differences in executive functions, emotional processing and decision-making [@barrett2015interoceptive; @petzschner2021computational; @murphy2019classifying]. Although little research interest has been conferred to investigating the relationship between interoception and deceptive ability *per se*, previous decision-making studies have demonstrated a negative correlation between interoceptive awareness, a metacognitive dimension of interoception, and one's likelihood to make risky decisions [@dunn2010listening; @furman2013interoceptive]. In fact, this is consistent with the somatic marker hypothesis, which posits an association between interoception and a propensity towards making rational decisions [@damasio1996somatic].

@@

The aim of the present study was to explore the contribution of ToM and interoception abilities on individuals' deceptive skills, as indicated by their lying confidence, physiological arousal and response time. To this end, we designed a directed-lying paradigm with 2 conditions differing in the nature of their feedback cues. The *Interrogation* condition was designed to emphasize (and preferentially mobilize) ToM-related mechanisms, whereas the *Polygraph* condition was designed to emphasize interoceptive mechanisms. We expected ToM and interoception to positively predict lying ability (i.e., higher lie confidence, shorter response time and lower physiological arousal), in particular in the *Polygraph* and the *Interrogation* condition, respectively.

@@

30 university students were recruited through posters, flyers, and online social media platforms. Inclusion criteria included residing in Singapore, and an absence of current or history of neurological, psychiatric, or cardiovascular disorders. Four participants were excluded as their data was not recorded due to technical issues. The final sample consists 26 participants (Mean age = 20.9, SD = 2.0, range:[18, 25], Sex: 65.4% women, 34.6% men). The heart rate of one participant and response time of one participant were excluded from further analysis due to extreme outlying values.

This pre-registered study (**OSF?** **Dom: I think so but not sure were to find the prereg**) was approved by the NTU Institutional Review Board (NTU IRB-2018-10-029)(**TBC**) and all procedures performed were in accordance with the ethical standards of the institutional board and with the 1964 Helsinki Declaration. All participants provided their informed consent prior to participation and were awarded with academic credits upon completion of the study.

@@

Two measures of ToM and related constructs were administered. The Yoni Task **[REF]** is a behavioral task in which participants have to **INSERT SHORT DESCRIPTION**. It captures 3 ToM dimensions, namely cognitive, physical and affective ToM abilities [@shamay2007dissociable].

The Basic Empathy Scale [BES, @jolliffe2006development], a 20-item questionnaire measuring two dimensions of empathy (cognitive and affective) using 5-point Likert scales was administered. Although ToM and empathy are regarded as distinct psychological constructs, previous research findings point to them being closely related [@gallant2020developing; @sebastian2012neural]. 
<!-- Specifically, empathy is often thought to be an integral component in the affective dimension of ToM (i.e., the ability to infer what someone else is feeling) [@shamay2010role].  -->

### Interoception

To assess participants' interoceptive ability, participants completed a Heartbeat Counting Task [HCT, @schandry1981heart] while having their actual heartbeats recorded. During the HCT task, participants were instructed to count the number of heartbeats over 5 trials with varying time intervals (20s, 25s, 30s, 35s, 40s, randomized order). Interoceptive accuracy was computed from the difference between the estimated number and the real number of heart beats. Interoceptive sensibility was estimated as the average of the confidence ratings presented at the end of each trial. Interoceptive awareness was indexed by the correlation between the objective accuracy and the subjective confidence.

The MAIA-2 [@mehling2012multidimensional], a 32-item questionnaire which measures 8 dimensions of interoception (Noticing, Not-Distracting, Not-Worrying, Attention, Regulation, Emotional Awareness, Self-Regulation, Body Listening, and Trust) using 5-point Likert scales, was also administered.

### Deception

**TODO: create a procedure figure**

We designed and implemented in PsychoPy [@peirce2019psychopy2] a directed-lying task in which the participants were instructed that they would have to answer 80 questions by either lying or telling the truth (depending on whether they see "lie" or "truth" written on the screen). Importantly, they were told that for half of the trials, they would have to convince another participant, located in another room, that would see them through a webcam connection (*Interrogation* condition). For the remaining trials, they would have to convince a "lie detection machine" that would be based on their physiological signals *Polygraph* condition. The two conditions were presented in a counter-balanced order, and each comprised of 40 trials.

The sequence of each trial was the same for both conditions. A question was shown on a computer screen, either phrased directly (e.g., "What is your favourite sport?"), or indirectly (e.g., "Is your favourite sport Hockey?"), with accompanying instructions below that either asked participants to verbally lie or tell the truth (half of  the trials in each condition, i.e., n = 20, were directed lies). Following a short interval (0.7 - 1.5s) for response preparation (during which was written "Connecting..." on the screen), participants were presented with either a social-feedback or biofeedback cue (for the *Interrogation* and *Polygraph* conditions, respectively), displayed for a maximum of 10s. Participants had to provide their answer to the question verbally (out loud) and press the space key to mark the end of their response. After another short interval (1.5 - 2.5s) during which was written "Disconnecting..." on the screen, participants were asked to rate n an analog scale  their confidence on how convincing (i.e., likely to get judged as truthful) they perceived their response to be.

In the *Interrogation* condition, participants had to answer each trial while receiving social feedback in the form of a video stream of an examiner, and were informed that the examiner would be evaluating the truthfulness of their responses by observing them through the live video feed. In actuality, the video was a pre-recorded video clips of a confederate's face (staying still with minimal reactions, and with a medical mask, as the experiment was run during COVID restrictions), and the same video-clips were used with all participants. Note that after debriefing, all participants reported believing that the stream was real and that the examiner was really there.

In the *Polygraph* condition, participants had to answer each trial while receiving biofeedback in the form of physiological signals (ECG, EDA and Respiration) being recorded. Participants were informed that they were being shown live feedback of their own physiological activity. In actuality, we displayed pre-recorded video clips of a confederate's signals. Note that after debriefing, all participants reported believing that it was indeed their own signals being shown.

Finally, on top of the deception task, we also measured the self-reported tendency to lie in their everyday life using the *Lie Scale* [@makowski2021structure], a 16-item questionnaire that assesses 4 dispositional lying dimensions (Ability, Negativity, Contextuality and Frequency) with visual analog scales.

<!-- ### Demographic Information -->

<!-- We further collected demographic information including age, ethnicity, nationality, handedness (for application of EDA electrodes), religion, duration lived in Singapore as well as highest education qualification achieved (or in the midst of achieving).  -->

@@

In the first phase of the study, participants answered a brief demographic survey before being administered a battery of questionnaires (*BES*, *MAIA-2*, *Lie Scale*) and cognitive-behavioural tasks (*HCT*, *Yoni Task*). This series of tasks and questionnaires were administered in a randomized order in order to control for order effects. In the second phase of the study, the *Deception Task* was administered.

Physiological recording devices were prepared prior to beginning the experimental paradigm. Using three ECG electrodes placed according to a modified Lead II configuration [@takuma1995alternative], cardiovascular activity, sampled at 1000 Hz, was recorded via the BioPac MP160 system (BioPac Systems Inc., USA).
